\batcave: Bayesian Analysis Tools for Context-Aware Variant Evaluation

### 2019-09-17 NAR_DRAFT.tex
extraneous and
constant not uniform

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is uniform across the genome.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.

>>>>

Many methods have been developed for calling somatic mutations from NGS data.
The earliest widely used somatic variant callers developed specifically for tumors, MuTect1 and Varscan2, used a combination of heuristic filtering and a model of sequencing errors to identify and score potential variants and set a threshold score designed to balance sensitivity and specificity \citep{Koboldt2012,Cibulskis2013}.
Subsequent research gave rise to a number of alternate strategies, including haplotype-based calling \citep{Garrison2012},
joint genotype analysis (SomaticSniper, JointSNVMix2, Seurat, and CaVEMan,MuClone) \cite{Larson2012,Roth2012a,Christoforides2013,Jones2016,Dorri2019}, allele frequency-based analysis (Strelka, MuTect, LoFreq, EBCall, deepSNV, LoLoPicker, and MuSE) \citep{Saunders2012,Wilm2012,Shiraishi2013b,Gerstung2012,Carrot-Zhang2017,Fan2016}, and ensemble and deep learning methods (MutationSeq, SomaticSeq, SNooPer, and BAYSIC) \rngcomment{Citations?}.
These methods vary in their complexity and specific focus.
But they all implicitly or explicitly assume that the rate of mutation is constant across the genome.

@@

\subsection{Estimation of the mutation profile}
Many aspects of genomic architecture can affect the somatic mutation rate at multiple scales \cite{Buisson2019}.
Here we focus on a small-scale feature, the tri-nucleotide context, which is known to strongly effect the prior probability of mutation \citep{Nik-Zainal2012a,Alexandrov2015,Lee-Six2018}.
The tri-nucleotide context of a genomic site consists of the identity of the reference base and the 3' and 5' flanking bases.
Folding the central base to the pyrimidines, there are two possible bases at the focal site, and there are four possible bases 3' and 5' of the focal site, yielding $2 \cdot 4 \cdot$ possible tri-nucleotide contexts $C$.
At the focal site, a mutation $m$ can be to any of three alternate alleles.
Indexing by the $c=\{1 \dots 32\}$ contexts and by the $m = \{1 \dots 3\}$ alternate bases, we have 96 possible substitution types $S_{m,c}$.
Eq.~\ref{eqn:full_prior} is then
\begin{equation}
  \label{eqn:detailed_prior}
  \begin{array}{rcl}
  \mathrm{p}(\mathrm{S}_{m,c}) &=&  \mathrm{p}(m, M | C = c) \\
                            &=& \mathrm{p}(m \mid C = c) \mathrm{p}(C = c \mid M)\frac{\mathrm{p}(M)}{\mathrm{p}(C = c)}.
  \end{array}
\end{equation}
The first two terms on the right-hand side can be estimated from the observed mutation profile (Figure ~\ref{NAR-sigfig}).
%Here $\mathrm{p}(\mathrm{S}_{m,c})$ represents probability that a substitution of type $\mathrm{S}_{m,c}$ will occur.
